Lossy compression of DNA sequencing quality data

### 2015-05-12 manuscript/q-compression.tex
more tweaks

\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded
from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
which are generated from a fragment library (insert size of 180 nt;
1,145,8940 paired-end reads). The corresponding reference sequence was
obtained from the \textsc{ncbi} RefSeq database (NC\_000913.2).

\textit{Mus musculus} data was downloaded from
\url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which
consisted of 18,828,274 reads of length 36.

@@

\begin{figure*}[!tb]
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods applied to four data sets:
  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponding to original
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

We compare the quality-value \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
different compressed representations, which include both fixed and
marginal costs (Figure \ref{fig:mse_vs_bpbp}). Here we present the
fragment libraries for the \textit{Rhodobacter sphaeroides}, and
\textit{Homo sapiens} data sets; the short-jump library results are
available in Supplemental Table 1. Storing the uncompressed quality
values requires 1 byte/base-call because they are stored in
\textsc{ascii} format, which is denoted by the dotted black asterisk
in the figure. The lossless compression of each data set using
\textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call and is denoted by
the colored asterisks on the abscissa; these values provide a
reference point using a widely available compression program. Values
from the same class of compression method tend to cluster together
across the different data sets, with the 0-degree polynomial
regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
lowest bits/base-call.

>>>>

\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded
from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
which are generated from a fragment library (insert size of 180 nt;
1,145,8940 paired-end reads; read length 300 nt). The corresponding
reference sequence was obtained from the \textsc{ncbi} RefSeq database
(NC\_000913.2).

\textit{Mus musculus} data was downloaded from
\url{http://trace.ddbj.nig.ac.jp/DRASearch/run?acc=SRR032209}, which
consisted of 18,828,274 reads of length 36 nt.

@@

\begin{figure*}[!tb]
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods and subsequently further
  compressed using a general algorithm via \textsc{bz}ip2 to four data
  sets: \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponding to original
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

We compare the quality-value \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
different compressed representations, which include both fixed and
marginal costs, and subsequently further compressed using a general
algorithm via \textsc{bz}ip2 (Figure \ref{fig:mse_vs_bpbp}). Here we
present the fragment libraries for the \textit{Rhodobacter
  sphaeroides}, and \textit{Homo sapiens} data sets; the short-jump
library results are available in Supplemental Table 1. Storing the
uncompressed quality values requires 1 byte/base-call because they are
stored in \textsc{ascii} format, and the lossless compression of each
data set using \textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call;
these values provide a reference point using a widely available
compression program. Values from the same class of compression method
tend to cluster together across the different data sets, with the
0-degree polynomial regression, profile encodings, and
\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.
